Overview and Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1)

### 2014-03-28 WSSSPE1_report/WSSSPE1_summary_report.tex
polishing sec8

% == Notes and ToDos ==
% 1. Why do we classify software cases into general purpose and special purpose? what is different? What are the sustainability challenges each face? Are they radically different? What kind of research do each category support? Are there separate questions to ask and answer for each category? Can we assume that the general purpose software has a wider scope and user community and challenges specific to these kinds of environments are not relavent in the special purpose scientific software which tend to be developed in the small?

% 2. What are the factors that has driven the development of these cases? How was the development started? What was the spark? Was there a design approach followed? Is the software modular/extensible? Can we identify unique and common factors and the reasons behind them.

% 3. Can we characterize the process itself for the selected cases: collaborative, seperate.

% 4. ToDo: summarize and report on an inclusive list (might not be exhaustive) of all the software products discussed in the workshop. (If possible color code the representaion according to a scheme based on the different dimensions pertaining to them.)

@@

% 6. Is there a conscious effort to deal with issues such as backward compatibility, ...

% 7. Career path: are the original developer(s) still around? Is most of the discussion around career path issues a myth after all?

@@

% 9. Has the project proved to be sustainable so far? Does the developers, managers make conscious efforts to make sure the project is sustainable? What determines the sustainability of the project in question: science, publications, usability, funding, reusability? Perhaps all factors to some extent? What factors are dominating? Can we send out a message to current scientific project to emphasize on one or more of these or other aspects?
% ====

For the purpose of studying software case studies, we classify the software
projects discussed in the workshop in two broad categories. First, utility
software, comprising software tools that enable and/or facilitate the
development of other tools and techniques to carry out scientific work. This
includes the software developed to efficiently utilize new research
infrastructures.  Second, scientific software, comprising software that was
developed with an aim to solve a class of scientific problems.

These categories have overlap in some cases where case studies do not cleanly
fit into a single category.  Our classification is based on two criteria:
first, the character of a software project that dominates and where it `seems'
to be the best fit, and, second, the workshop discussion session where the
paper about the project was discussed.

Our discussion is aimed at understanding the practicalities of the points
discussed in the workshop, and we hope to extract the best practices and
associate them with the projects. We discuss how the cases are applicable to
and have impacted user communities over their lifetime, isolating similarities
and differences between projects. We discuss the takeaways for a new
project and the lessons learned. We try to understand how the course of current
and future development projects might be altered to implement lessons learned.

\subsection{Utility Software}
Utility software has the potential of the broadest community impact of the two
categories. This is especially relevant in the modern era of social computing.
In the context of research, a utility software project could be defined as
general purpose software applicable to one or many generic tasks and/or
enabling other software to run by providing a suitable environment. Examples
are collaborative development frameworks such as \emph{GitHub} and
\emph{Bitbucket}, workflow and scripting frameworks such as \emph{Galaxy},
\emph{Swift}, \emph{Globus} and \emph{VisTrails}, and visualization frameworks
such as \emph{VisIT}. Often projects start with a specific, small set of
requirements and users in mind but expand to provide more generic functionality
and appeal to a wider user community. Owing to its generic nature, utility
software finds a wider user community which in turn leads to a higher
participation in development. Consequently, the process becomes user-driven and
self-sustaining.

@@

Utility software is often developed aimed at better utilizing particular, new
infrastructures and architectures, e.g., \emph{MVAPICH}, \emph{VisIT}.
Development in many cases is often a high risk/reward undertaking and explores
uncharted territories.
%
%\note{Dan: so is the only difference here that this category is aimed at new systems, where utility software is not?  If so, perhaps this should be incorporated into that subsection, so there would just be two subsections.  Something in the framing of this subsection doesn't feel right to me.  FL: I respectfully disagree the the only reason to develop frameworks is to better utilize (new or old) infrastructure. Another really important reason is to modularize development, making it much easier for an individual developer/user to integrate their own code. Ketan: I think in this section I wanted to bring up a class of software projects which are undertaken with the main aim of tackling new systems and architectures in mind. New infrastructures such as GPU, Cloud, fast IB, federal science networks, etc. were on my mind as I see many submissions around them in the workshop.}
%
Successful projects reap high rewards and have longer usage span. Long
requirement-development-test cycles.  Wide community support but tends to be
handled by specialists rather then end users. Promotes collaborations across
the breadth (e.g., different departments) and depth (e.g., stakeholders within
a department) of community, one of the key ingredient of a sustainable process.
Ferenbaugh~\cite{Ferenbaugh_WSSSPE} discusses the risks and challenges
associated with software projects dealing with new architectures. To leverage
the power of accelerators such as GPUs and MICs new code and libraries are
required. The experience of efforts as described in~\cite{Ferenbaugh_WSSSPE}
met with a limited success but nonetheless with many invaluable lessons were
learned about influential cultural and technical aspects in sustainable
software development practices.

>>>>

% == Notes and ToDos ==
% 1. Why do we classify software cases into general purpose and special
% purpose? what is different? What are the sustainability challenges each face?
% Are they radically different? What kind of research do each category support?
% Are there separate questions to ask and answer for each category? Can we
% assume that the general purpose software has a wider scope and user community
% and challenges specific to these kinds of environments are not relavent in
% the special purpose scientific software which tend to be developed in the
% small?

% 2. What are the factors that has driven the development of these cases? How
% was the development started? What was the spark? Was there a design approach
% followed? Is the software modular/extensible? Can we identify unique and
% common factors and the reasons behind them.

% 3. Can we characterize the process itself for the selected cases:
% collaborative, seperate.

% 4. ToDo: summarize and report on an inclusive list (might not be exhaustive)
% of all the software products discussed in the workshop. (If possible color
% code the representaion according to a scheme based on the different
% dimensions pertaining to them.)

@@

% 6. Is there a conscious effort to deal with issues such as backward
% compatibility, ...

% 7. Career path: are the original developer(s) still around? Is most of the
% discussion around career path issues a myth after all?

@@

% 9. Has the project proved to be sustainable so far? Does the developers,
% managers make conscious efforts to make sure the project is sustainable? What
% determines the sustainability of the project in question: science,
% publications, usability, funding, reusability? Perhaps all factors to some
% extent? What factors are dominating? Can we send out a message to current
% scientific project to emphasize on one or more of these or other aspects?
% ====

% TODO: Why are we doing case studies?
Many important issues related to scientific software sustainability were
discussed in the previous sections. In this section we discuss some of the
software projects as case-studies to better understand the relevance of those
issues and how are they connected to the software sustainability and related
factors.

We classify the software projects discussed in the workshop in two broad
categories. First, the utility software, comprising of software that enable
and/or facilitate the development of other tools and techniques to carry out
scientific work. This includes the software developed to efficiently utilize
new research infrastructures. Second, scientific software, comprising software
that was developed with an aim to solve a specific scientific problem.

% Why are we classifying software like above?
This classification is motivated by our argument that the software projects
wildly vary in factors such as scope, purpose and usage. The development and
management of software is significantly influenced by these factors.
Consequently, the sustainability challenges faced by software differ and must
be treated separately. For instance, the challenges faced by a gateway software
development project such as \textit{CIPRES} are distinct to a niche software
for ab initio modeling and simulation such as \textit{VASP}.

% Below: What are the goals of this discussion? TODO: Better motivate. May be make a bried list of items discussed in the workshop.
The discussion is aimed at understanding the practicalities of the key points
discussed in the workshop, and we hope to extract the best practices and
associated with the case studies. We discuss how the cases are applicable to
and have impacted user communities over their lifetime, isolating similarities
and differences between projects. We discuss the takeaways for new
projects and the lessons learned. We try to understand how the course of current
and future development projects might be altered to implement lessons learned.

\subsection{Utility Software}
Utility software has the potential of a broader user base. This is especially
relevant in the modern era of social computing. In the context of research, a
utility software project could be defined as general purpose system applicable
to one or many generic tasks and/or enabling other software to run by providing
a suitable environment. Examples are collaborative development frameworks such
as \emph{GitHub} and \emph{Bitbucket}, distributed and workflow computing
frameworks such as \emph{Galaxy}, \emph{Swift}, \emph{Globus} and
\emph{VisTrails}, and visualization frameworks such as \emph{VisIT}. Often
projects start with a specific, small set of requirements and users in mind but
expand to provide functionality and appeal to a wider user community. Owing to
its generic nature, utility software finds a wider user community which in turn
leads to a higher participation in development.  Consequently, the process
becomes user-driven and self-sustaining.

@@

Utility software is often developed aimed at better utilizing particular, new
infrastructures and architectures, e.g., \emph{MVAPICH}, \emph{VisIT}.
Development in many cases is often a high risk/reward undertaking and explores
uncharted territories.
%
%
%\note{Dan: so is the only difference here that this category is aimed at new systems, where utility software is not?  If so, perhaps this should be incorporated into that subsection, so there would just be two subsections.  Something in the framing of this subsection doesn't feel right to me.  FL: I respectfully disagree the the only reason to develop frameworks is to better utilize (new or old) infrastructure. Another really important reason is to modularize development, making it much easier for an individual developer/user to integrate their own code. Ketan: I think in this section I wanted to bring up a class of software projects which are undertaken with the main aim of tackling new systems and architectures in mind. New infrastructures such as GPU, Cloud, fast IB, federal science networks, etc. were on my mind as I see many submissions around them in the workshop.}
%
%
Successful projects reap high rewards and have longer usage span. Long
requirement-development-test cycles.  Wide community support but tends to be
handled by specialists rather then end users. Promotes collaborations across
the breadth (e.g., different departments) and depth (e.g., stakeholders within
a department) of community, one of the key ingredient of a sustainable process.
Ferenbaugh~\cite{Ferenbaugh_WSSSPE} discusses the risks and challenges
associated with software projects dealing with new architectures. To leverage
the power of accelerators such as GPUs and MICs new code and libraries are
required. The experience of efforts as described in~\cite{Ferenbaugh_WSSSPE}
met with a limited success but nonetheless with many invaluable lessons were
learned about influential cultural and technical aspects in sustainable
software development practices.
